Calibrating seed-based mapping heuristics with Sesame

### 2020-03-25 manuscript.tex
Add section about realistic error models

Spaced seeds of weight $w$ (the number of 1s in the model) have the same
memory requirements as continuous seeds of length $w$ but they are more
sensitive~\cite{li2006superiority}, making them very attractive for
homology search. They were used in the first generation of short-read
mappers~\cite{pmid18684737, pmid19461883, pmid19675096, pmid2945724}, but
nowadays they find more applications in genome
comparisons~\cite{pmid21209072, healy2016flak},
metagenomics~\cite{pmid26209798, pmid27540266}, genome
assembly~\cite{pmid26539459} and long-read mapping~\cite{pmid27079541}.

The success of the mainstream mappers BWA-MEM~\cite{li2013aligning} and
Bowtie2~\cite{pmid22388286} is due in part to the FM-index, which only
supports continuous seeds. Some workarounds are
available~\cite{horton2008dislex, gagie2017compressed} but they increase
the memory footprint, explaining that short reads are typically mapped
using continuous seeds. More generally, computing the sensitivity of
spaced seeds is challenging~\cite{pmid16819802, li2006superiority,
martin2017faster}. It is theoreticlally possible to do this using the
tools introduced below, but the complexity becomes prohibitive for seed
models with several don't-care positions. For these reasons we will not
treat the case of spaced seeds and we will give more detail in
section~\ref{sec_disc} bout the kind of difficulties they represent for
our approach.

@@

The analytic combinatorics framework relies on a strategy referred to as
the \emph{symbolic method}~\cite{sedgewick2013introduction}. The idea is
to combine simple objects into more complex objects. Each combinatorial
operation on the objects corresponds to a mathematical operation on their
weighted generating functions. One can thus obtain the weighted generating
function of complex objects, whose coefficients $a_k$ are the
probabilities of interest.

As explained in~\cite{filion2017analytic} and~\cite{filion2018analytic},
we recode the reads in alphabets of custom symbols and we specify a
construction plan of the reads using a \emph{transfer matrix} $M(z)$. The
transfer matrix specifies which types of segments can follow each other in
the reads of interest: the entry at coordinates $(i,j)$ is the weighted
generating function of segments of type $j$ that can be appended to
segments of type $i$.

@@

\begin{table}
% MEM seeds.
\renewcommand{\arraystretch}{1.1}
\centering
\begin{tabular}{ccccc}
k & N & Not seeded & Seeded, not best & \% seed failure \\
\hline
50  &  1  & $7.8 \cdot 10^{-4}$ & $9.5 \cdot 10^{-5}$ & $89.1\%$ \\
50  & 10  & $7.8 \cdot 10^{-3}$ & $9.1 \cdot 10^{-4}$ & $89.6\%$ \\
50  & 100 & $7.6 \cdot 10^{-2}$ & $5.7 \cdot 10^{-3}$ & $93.0\%$ \\
100 &  1  & $2.1 \cdot 10^{-5}$ & $7.6 \cdot 10^{-6}$ & $73.7\%$ \\
100 & 10  & $2.6 \cdot 10^{-4}$ & $7.1 \cdot 10^{-5}$ & $78.1\%$ \\
100 & 100 & $5.3 \cdot 10^{-3}$ & $5.9 \cdot 10^{-4}$ & $90.1\%$
\end{tabular}
\caption{Numerical simulating of mapping errors with MEM seeds. Reads are
mapped incorrectly if either the true location is not in the candidate set
(not seeded) or it is in the candidate set but is not the best location
(seeded, not best). Ideal reads were randomly generated 1,000,000 times
with sequencing error rate $p = 0.01$, divergence rate $\mu = 0.06$,
varying the read size $k$ and number $N$ of duplicates as indicated in the
table. Mapping errors of both types were recorded, and the fraction of the
first to the total was computed.}
\label{table_mem}
\end{table}

@@

\begin{table}
% Exact seeds.
\renewcommand{\arraystretch}{1.1}
\centering
\begin{tabular}{ccccc}
k & N & Prob. seeding error & Prob. other error & \% seeding \\
\hline
50  &  1  & $3.3 \cdot 10^{-5}$  & $5.1 \cdot 10^{-4}$ & $6.0\%$ \\
50  & 10  & $2.9 \cdot 10^{-4}$  & $5.0 \cdot 10^{-3}$ & $5.4\%$ \\
50  & 100 & $1.0 \cdot 10^{-3}$  & $4.6 \cdot 10^{-2}$ & $2.2\%$ \\
100 &  1  & $1.5 \cdot 10^{-10}$ & $2.3 \cdot 10^{-5}$ & $< 0.1\%$ \\
100 & 10  & $1.2 \cdot 10^{-9}$  & $2.3 \cdot 10^{-4}$ & $< 0.1\%$ \\
100 & 100 & $2.6 \cdot 10^{-9}$  & $2.3 \cdot 10^{-3}$ & $< 0.1\%$
\end{tabular}
\caption{My least favorite table.}
\label{table_exact}
\end{table}

\begin{table}
% Skip-9 seeds.
\renewcommand{\arraystretch}{1.1}
\centering
\begin{tabular}{ccccc}
k & N & Prob. seeding error & Prob. other error & \% seeding \\
\hline
50  &  1  & $1.3 \cdot 10^{-4}$ & $4.9 \cdot 10^{-4}$ & $20.1\%$ \\
50  & 10  & $1.2 \cdot 10^{-3}$ & $4.8 \cdot 10^{-3}$ & $19.7\%$ \\
50  & 100 & $5.5 \cdot 10^{-3}$ & $4.5 \cdot 10^{-2}$ & $10.9\%$ \\
100 &  1  & $2.0 \cdot 10^{-8}$ & $2.3 \cdot 10^{-5}$ & $< 0.1\%$ \\
100 & 10  & $1.5 \cdot 10^{-7}$ & $2.3 \cdot 10^{-4}$ & $< 0.1\%$ \\
100 & 100 & $3.9 \cdot 10^{-7}$ & $2.3 \cdot 10^{-3}$ & $< 0.1\%$
\end{tabular}
\caption{My not so favorite table.}
\label{table_skip}
\end{table}

@@

@@

\subsection{Realistic error patterns}

This section will come later.

@@

>>>>

Spaced seeds of weight $w$ (the number of 1s in the model) have the same
memory requirements as contiguous seeds of length $w$ but they are more
sensitive~\cite{li2006superiority}, making them very attractive for
homology search. They were used in the first generation of short-read
mappers~\cite{pmid18684737, pmid19461883, pmid19675096, pmid2945724}, but
nowadays they find more applications in genome
comparisons~\cite{pmid21209072, healy2016flak},
metagenomics~\cite{pmid26209798, pmid27540266}, genome
assembly~\cite{pmid26539459} and long-read mapping~\cite{pmid27079541}.

The success of the mainstream mappers BWA-MEM~\cite{li2013aligning} and
Bowtie2~\cite{pmid22388286} is due in part to the FM-index, which only
supports contiguous seeds. Some workarounds are
available~\cite{horton2008dislex, gagie2017compressed} but they increase
the memory footprint, explaining that short reads are typically mapped
using contiguous seeds. More generally, computing the sensitivity of
spaced seeds is challenging~\cite{pmid16819802, li2006superiority,
martin2017faster}. It is theoreticlally possible to do this using the
tools introduced below, but the complexity becomes prohibitive for seed
models with several don't-care positions. For these reasons we will not
treat the case of spaced seeds and we will give more detail in
section~\ref{sec_disc} bout the kind of difficulties they represent for
our approach.

@@

The analytic combinatorics framework relies on a strategy referred to as
the \emph{symbolic method}~\cite{sedgewick2013introduction}. The idea is
to combine simple objects into more complex objects. Each combinatorial
operation on the objects corresponds to a mathematical operation on their
weighted generating functions. One can thus obtain the weighted generating
function of complex objects, whose coefficients $a_k$ are the
probabilities of interest~\cite{regnier2000unified, nicodeme2002motif,
sedgewick2013introduction}.

As explained in~\cite{filion2017analytic} and~\cite{filion2018analytic},
we recode the reads in alphabets of custom symbols and we specify a
construction plan of the reads using a so-called \emph{transfer matrix}
$M(z)$. The transfer matrix specifies which types of segments can follow
each other in the reads of interest: the entry at coordinates $(i,j)$ is
the weighted generating function of segments of type $j$ that can be
appended to segments of type $i$.

@@

\begin{table}
% MEM seeds.
\renewcommand{\arraystretch}{1.1}
\centering
\begin{tabular}{ccccc}
k & N & Not seeded & Seeded, not best & \% seed failure \\
\hline
50  &  1  & $9.2 \cdot 10^{-4}$ & $8.4 \cdot 10^{-5}$ & $91.7\%$ \\
50  & 10  & $9.0 \cdot 10^{-3}$ & $7.9 \cdot 10^{-4}$ & $92.0\%$ \\
50  & 100 & $8.1 \cdot 10^{-2}$ & $5.1 \cdot 10^{-3}$ & $94.1\%$ \\
100 &  1  & $2.4 \cdot 10^{-5}$ & $7.3 \cdot 10^{-6}$ & $76.7\%$ \\
100 & 10  & $2.9 \cdot 10^{-4}$ & $6.9 \cdot 10^{-5}$ & $80.1\%$ \\
100 & 100 & $5.5 \cdot 10^{-3}$ & $5.7 \cdot 10^{-4}$ & $90.7\%$
\end{tabular}
\caption{Numerical simulation of mapping errors with MEM seeds. Reads are
mapped incorrectly if either the true location is not in the candidate set
(not seeded) or it is in the candidate set but is not the best location
(seeded, not best). Ideal reads were randomly generated 1,000,000 times
with sequencing error rate $p = 0.01$, divergence rate $\mu = 0.06$,
varying the read size $k$ and number $N$ of duplicates as indicated in the
table and using MEM seed of minimum size $\gamma = 19$. Mapping errors of
both types were recorded, and the fraction of the first to the total was
computed.}
\label{table_mem}
\end{table}

@@

\begin{table}
% Skip-9 seeds.
\renewcommand{\arraystretch}{1.1}
\centering
\begin{tabular}{ccccc}
k & N & Prob. seeding error & Prob. other error & \% seeding \\
\hline
50  &  1  & $1.7 \cdot 10^{-4}$ & $4.9 \cdot 10^{-4}$ & $25.4\%$ \\
50  & 10  & $1.6 \cdot 10^{-3}$ & $4.7 \cdot 10^{-3}$ & $24.9\%$ \\
50  & 100 & $7.9 \cdot 10^{-3}$ & $4.4 \cdot 10^{-2}$ & $15.2\%$ \\
100 &  1  & $4.4 \cdot 10^{-8}$ & $2.3 \cdot 10^{-5}$ & $0.2\%$ \\
100 & 10  & $3.5 \cdot 10^{-7}$ & $2.3 \cdot 10^{-4}$ & $0.2\%$ \\
100 & 100 & $1.0 \cdot 10^{-6}$ & $2.3 \cdot 10^{-3}$ & $< 0.1\%$
\end{tabular}
\caption{Numerical simulation of mapping errors with skip-9 seeds. Reads
are mapped incorrectly if either the true location is not in the candidate
set (not seeded) or it is in the candidate set but is not the best
location (seeded, not best). Ideal reads were randomly generated 1,000,000
times with sequencing error rate $p = 0.01$, divergence rate $\mu = 0.06$,
varying the read size $k$ and number $N$ of duplicates as indicated in the
table and using skip-9 seed of minimum size $\gamma = 19$. Mapping errors
of both types were recorded, and the fraction of the first to the total
was computed.}
\label{table_skip}
\end{table}

@@

\subsection{More realistic error models}

In order to make the model tractable, we had to make some simplifying
assumptions regarding the distribution of errors
(section~\ref{sec:error}). In particular, we assumed that sequencing
errors are uniform on the read, which is known to not be the case with
current instruments. The errors are typically more frequent at the ends of
the reads, meaning that seeds may be longer than expected, impacting the
probability that seeding is off target.

To test whether this is the case, we used the ART
simulator~\cite{pmid22199392} to emulate the error distribution of
Illumina reads as per technology standards of 2016. We used the settings
for the Illumina HiSeq 2000 and generated random reads of size 50 and 100
from the human genome. We estimated the probability that seeding is off
target from simulations and compared the results to calculations performed
by Sesame (ART suggests error rates $p=0.0052$ and $p=0.0075$ for reads of
size 50 and 100, respectively).

Table~\ref{table_MEM_Illumina} shows the results for MEM seeds of minimum
size $\gamma=19$. The estimates show some discrepancy but Sesame is
accurate to within 15\% in these conditions.
Table~\ref{table_skip_Illumina} shows the same type of comparison for
skip-9 seeds of size $\gamma=19$. The values are substantially lower than
for MEM seeds because skip-9 seeds are more sensitive. In these
conditions, Sesame is accurate to within 35\% but it is important to
highlight that simulations are unreliable for events with very low
probability.

\begin{table}
% MEM seeds.
\renewcommand{\arraystretch}{1.1}
\centering
\begin{tabular}{cccc}
k & N & Simulation & Sesame \\
\hline
50  &  1  & $3.8 \cdot 10^{-4}$ & $4.5 \cdot 10^{-4}$ \\
50  & 10  & $4.4 \cdot 10^{-3}$ & $4.5 \cdot 10^{-3}$ \\
50  & 100 & $4.2 \cdot 10^{-2}$ & $4.0 \cdot 10^{-2}$ \\
100 &  1  & $3.5 \cdot 10^{-5}$ & $3.7 \cdot 10^{-5}$ \\
100 & 10  & $4.1 \cdot 10^{-4}$ & $4.2 \cdot 10^{-4}$ \\
100 & 100 & $8.6 \cdot 10^{-3}$ & $8.9 \cdot 10^{-3}$ 
\end{tabular}
\caption{Numerical simulation of ``seeding'' errors with MEM seeds.
Realistic reads were randomly generated 1,000,000 times with sequencing
error rate $p = 0.01$, divergence rate $\mu = 0.06$, varying the read size
$k$ and number $N$ of duplicates as indicated in the table and using
skip-9 seed of minimum size $\gamma = 19$. Mapping errors of both types
were recorded, and the fraction of the first to the total was computed.}
\label{table_MEM_Illumina}
\end{table}

\begin{table}
% Skip seeds.
\renewcommand{\arraystretch}{1.1}
\centering
\begin{tabular}{cccc}
k & N & Simulation & Sesame \\
\hline
50  &  1  & $4.5 \cdot 10^{-5}$ & $4.5 \cdot 10^{-5}$ \\
50  & 10  & $3.9 \cdot 10^{-4}$ & $4.2 \cdot 10^{-4}$ \\
50  & 100 & $1.9 \cdot 10^{-3}$ & $2.1 \cdot 10^{-3}$ \\
100 &  1  & $< 10^{-6}$         & $< 10^{-6}$         \\
100 & 10  & $2 \cdot 10^{-6}$   & $3 \cdot 10^{-5}$ \\
100 & 100 & $6 \cdot 10^{-6}$   & $8 \cdot 10^{-4}$ 
\end{tabular}
\caption{Numerical simulation of ``seeding'' errors with skip-9 seeds.
Realistic reads were randomly generated 1,000,000 times with sequencing
error rate $p = 0.01$, divergence rate $\mu = 0.06$, varying the read size
$k$ and number $N$ of duplicates as indicated in the table and using
skip-9 seed of minimum size $\gamma = 19$. Mapping errors of both types
were recorded, and the fraction of the first to the total was computed.}
\label{table_skip_Illumina}
\end{table}

Overall, these results suggest that the relatively strong assumptions
regarding the distribution of errors are not a major issue in practice.
The answer of course depends on the type of sequencer that is used and on
the particulars of the mapping problem.

In practical applications, it is likely that other factors will be more
detrimental for the precision of the estimates. For instance, the number
of duplicates $N$ and their divergence rate $\mu$ are unknown, introducing
an uncertainty that propagates to the Sesame estimates. Importantly, we
assumed that duplicates evolve independently of each other and with
uniform substitutions even in the simulations featured in
Table~\ref{table_MEM_Illumina} and \ref{table_skip_Illumina} because more
realistic models are tedious to implement. The burden of this assumption
on the precision of the estimate in practical settings is difficult to
gauge. Overall, Sesame estimates should be considered approximate in all
applications with real biological data.

@@

@@

We have not treated the case of spaced seeds here. It is possible to find
a transfer matrix for reads that contain no match for a given spaced seed,
but as the number of don't-care positions increases, the dimensions of the
matrix become prohibitively large. This is a well known difficulty. The
analytic combinatorics approach presented here is equivalent to defining
finite automata that recognize seed-containing reads. Automata can be used
to compute the sensitivity of spaced seeds~\cite{kucherov2005multiseed,
buhler2005designing}, with the caveat that such automata can contain a
very large amount of states. Overall, the transfer matrix method does not
provide any significant advantage over existing alternatives for spaced
seeds.
